Racial Force Probe

### 2024-02-15 paper/paper-draft.qmd

---
title: "Prison Health"
author:
  - Monica Sainani
  - Kaavya Kalani
  - Aamishi Avarsekar
thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
date: today
header-includes:
    - \usepackage{setspace}\doublespacing
date-format: long
abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations. The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being. Utilizing data collected over the past 15 years, this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners. By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being. This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities. Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016. For our analysis we will not be considering 2001, but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over.

@@

# Results
Based on the original paper’s findings, we decided to further investigate the health and general well-being of prisoners and trends of diseases within prison, causes of death and also the trends of COVID-19 across a 7-day period.

@@

# Create the plot
ggplot(health_data_by_year_long_prop, aes(x = as.numeric(Year), y = Proportion, color = Disease)) +
  geom_line(linewidth = 1.0) +
  labs(title = "Trends of Diseases by Year",
       x = "Year",
       y = "Proportion of cases",
       color = "Disease") +
  theme_minimal()
```

#Figure 1 
compares shows the cause of death among prisoners due to diseases. In order to accurately investigate, we adjusted the number of deaths caused by a particular disease by the total number of prisoners recorded for that year. This accounts for the differences in the number of deaths recorded between each cause.

@@

# Create the figure using ggplot2 with solid lines
ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
  geom_line(linewidth = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
  scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases") +
  theme_minimal() +
  geom_vline(xintercept = as.numeric(as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01"))),
             linetype = "dashed", color = "gray") +
  annotate("text", x = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01")),
           y = 150, label = c("April 2020", "May 2020", "June 2020", "July 2020", "August 2020", "September 2020"),
          vjust = -0.5, hjust = 0, color = "gray") +
  scale_color_manual(values = c("blue", "red", "green"))

#figure 2

@@

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-race-2
#| fig-cap: Causes of deaths among the prisoners by race

@@

>>>>

---
title: "Racial Force Probe"
author:
  - Monica Sainani
  - Kaavya Kalani
  - Aamishi Avarsekar
thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
date: today
date-format: long
abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations. The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being. Utilizing data collected over the past [insert number of years], this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners. By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being. This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities. Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

msfp0116stt02data.csv contains counts for causes of deaths of the prisoners for the years 2001, 2006-2016. For our analysis we will not be considering 2001 but only 2006 to 2016 for the trends as it gives us a continuous period to observe trends over.

@@

# Results

Based on the original paper’s findings, we decided to further investigate the health and general well-being of prisoners and trends of diseases within prison, causes of death and also the trends of COVID-19 across a 7-day period.

@@

# Create the plot
ggplot(health_data_by_year_long_prop, aes(x = as.numeric(Year), y = Proportion, color = Disease)) +
  geom_line(size = 1.0) +
  labs(title = "Trends of Diseases by Year",
       x = "Year",
       y = "Proportion of cases",
       color = "Disease") +
  theme_minimal()

```
Figure 1 compares shows the cause of death among prisoners due to diseases. In order to accurately investigate, we adjusted the number of deaths caused by a particular disease by the total number of prisoners recorded for that year. This accounts for the differences in the number of deaths recorded between each cause.

@@

# Create the figure using ggplot2 with solid lines
ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
  geom_line(size = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
  scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases per 100,000 Among
#Those in Prison, Prison Staff, and in the General Population: May to September
#2020") +
  theme_minimal() +
  geom_vline(xintercept = as.numeric(as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01"))),
             linetype = "dashed", color = "gray") +
  annotate("text", x = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01")),
           y = 150, label = c("April 2020", "May 2020", "June 2020", "July 2020", "August 2020", "September 2020"),
          vjust = -0.5, hjust = 0, color = "gray") +
  scale_color_manual(values = c("blue", "red", "green"))

```
result 2

@@

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-race
#| fig-cap: Causes of deaths among the prisoners by race

@@

# Load cleaned data
ts_data <- read.csv("data/analysis_data/cleaned_covid19_data.csv", header = TRUE, stringsAsFactors = FALSE)

# Create the figure using ggplot2 with solid lines
ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
  geom_line(size = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
  scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases per 100,000 Among
Those in Prison, Prison Staff, and in the General Population: May to September
2020") +
  theme_minimal() +
  geom_vline(xintercept = as.numeric(as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01"))),
             linetype = "dashed", color = "gray") +
  annotate("text", x = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01")),
           y = 150, label = c("April 2020", "May 2020", "June 2020", "July 2020", "August 2020", "September 2020"),
           vjust = -0.5, hjust = 0, color = "gray") +
  scale_color_manual(values = c("blue", "red", "green"))

# Load cleaned data
health_demographic_data_clean <- read_csv("data/analysis_data/cleaned_health_data_gender.csv")

# Plot the proportions graph using geom_bar with position = "dodge"
ggplot(health_demographic_data_clean, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +
  labs(title = "Proportions of Causes of Death by Category",
       x = "Gender",
       y = "Proportion") +
  scale_fill_discrete(labels = cause_mapping) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-race
#| fig-cap: Causes of deaths among the prisoners by race

# Load cleaned data from CSV
race_health_data_long <- read.csv("data/analysis_data/cleaned_health_data_race.csv", header = TRUE, stringsAsFactors = FALSE)
